{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2017-10-27 sftrends_manu.Rnw
minor edits to maybe final draft

\section*{Abstract}
\noindent Quantitative descriptions of water quality, physical, and biological characteristics in estuaries is critical for developing an ecological understanding of drivers of change.  Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \ac{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.  Trend analysis with flow-normalized results demonstrated the potential to interpret different changes relative to observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed.  We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \ac{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \ac{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and influenced by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

Understanding drivers of water quality changes in large estuaries depends on accurate descriptions of source inputs from the watershed. The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows in the upper \ac{sfe} that receives and processes inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}. Sediment transport and \ac{wwtp} inputs from the Delta are primary sources of nutrients for the estuary \citep{Dugdale07,Cornwell14}.  Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, changes in response to stressors suggests that recent conditions have not followed past trajectories. Increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms (e.g., \textit{Pseudo-nitzschia}, \textit{Alexandrium}, \textit{Dinophysis spp.}) have been a recent concern for the management of this prominent system \citep{Lehman05,Cloern05,Cloern07,Shellenbarger08}.  As a result, \ac{chla} thresholds to assess and manage levels of concern for phytoplankton biomass in the lower estuary have been proposed \citep{Sutula17}. Although these changes are linked to drivers at different spatial and temporal scales, describing inputs from the Delta is critical to understand downstream effects.

Rates of primary production in coastal habitats are often defined by nutrient concentrations, although a simple relationship between enrichment and water quality changes can be difficult to determine \citep{Cloern01}. Nutrient concentrations are generally non-limiting for phytoplankton growth in the upper \ac{sfe}.  In contrast, light conditions have long been considered the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}, whereas grazing pressure from pelagic fishes and benthic invertebrates can reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.  Moreover, changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.  \citet{Glibert14} attributed recent phytoplankton blooms in Suisun Bay to a drought, during which residence times and nitrification rates increased.  Speciation changes in the dominant forms of nitrogen are considered key factors that contribute to phytoplankton blooms, particularly seasonal reductions in ammonium that allow uptake of nitrate that stimulates growth \citep{Dortch90,Dugdale07}. Changes in phytoplankton biomass and the effects on pelagic organisms have also been of concern.  \citet{Jassby08} described decadal trends in phytoplankton biomass to understand mechanisms of decline for pelagic fish populations in the Delta.  Although phytoplankon concentrations have been relatively consistent in Suisan Bay, biomass in the upper Delta has been increasing.  Much of these trends were explained by invasion of benthic grazers in polyhaline areas and changes in the mean flow conditions observed in the Delta.

Long-term monitoring data are powerful sources of information that can facilitate descriptions of water quality change in coastal regions. A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep[\cref{fig:delt_map},][]{IEP13}.  Although these data have been used extensively, water quality trends covering the full spatial and temporal coverage of the monitoring dataset have not been systematically evaluated. Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.  An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported to the estuary \citep{Novick15}. Seasonal and annual changes in the delivery of water inflows and water exports directly from the system can also obscure trends \citep{Jassby00,Jassby08}.

@@

The goal of this study was to provide a comprehensive description of nutrient trends in the Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change. This study also supports a general need to more effectively use long-term monitoring data to inform management priorities for water quality in coastal regions.  The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
\item provide detailed descriptions of two case studies in the context of conceptual relationships modeled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.  Our general hypothesis was that the results were expected to support previous descriptions of trends in this well-studied system, but that new insight into spatial and temporal variation in response endpoints was expected, particuarly in flow-normalized model predictions.

@@

Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.  Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \citep{Cloern96,Thompson08}, limits on phytoplankton growth from high turbidity, and filter-feeding by bivalve mollusks in the northern portion \citep{Thompson08,Crauder16}.  However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales \citep[e.g., variation in sea surface temperatures,][]{Cloern07}.

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}. Although many other stations are available for the region, the stations were chosen because they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}. Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were below the deteciton limit (left-censored), although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal trend ($\beta_3$, $\beta_4$), and $Q$ is the flow variable (either flow or salinity depending on station).  Generally, the \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from \Sexpr{minobs} at \Sexpr{minsta} to \Sexpr{maxobs} at \Sexpr{maxsta}).  Rather than fitting a global model to the entire time series, one regression is fit to every observation.  Observations within a window for each regression are weighted relative to annual, seasonal, and flow distances from the observation at the center of the window.  Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.  This approach allows for a type of smoothing where the observed fit is specific to the data characteristics within windows. Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.  All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data below the detection limit \citep{Tobin58}.  Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September). All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.  The default model fitting procedures were used that set half-window widths as six months for seasonal weights, ten years for annual weights, and half the range of salinity or flow in the input data for $Q$ weights.

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows \citep{Hirsch10}. Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods. Flow-normalization predictions for each month of each year were based on the average of predictions for flow values that occur in the same month across all years, weighted within each specific month and year for every observation. Flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping. Annual groupings were chosen as approximate twenty year midpoints in the time series and seasonal groupings were chosen to evaluate inter-annual changes while keeping season constant.

@@

\subsection{Observed and modelled Data}

@@

Long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).  Although each station varied in the overall concentrations, patterns within the three Deltra regions (peripheral, interior, and Suisun) were observed.  Concentrations for all nitrogen analytes were highest in the peripheral stations.  Ammonium concentrations at P8 and C3 were highest and showed a consistent increase over time, followed by a reduction beginning in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.  By contrast, \ac{din} and nitrite/nitrate concentrations at the peripheral stations showed increases at  P8 and C10 followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did not show any noticeable trends.  Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for D26.  Trends in \ac{din} and nitrite/nitrate for the interior stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.  These trends were similar for the Suisun stations, although the reduction in the late 1990s did not occur.  By contrast, ammonium concentrations were low in Suisun but a gradual increase over the period of record was observed.

@@

@
A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends. For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant ($p > 0.05$) with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant. This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an increase in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positive or less negative) compared to observed trends.  \Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change, \Sexpr{eng(chgtb['zerotoneg'])} sites went from no change to negative estimated change, and \Sexpr{eng(chgtb['zerotopos'])} site went from no change to a positive trend for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}). The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.

@@

Water quality conditions in the Delta are dynamic and not easily characterized from observed time series.  Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions of the Delta that is not possible with raw observations.  A general conclusion is that nitrogen concentrations have decreased overall throughout the approximate forty years of observation.  These results are confirmed visually from \ac{wrtds} (\cref{fig:prddat}) and through significant results from trend tests (\cref{fig:trndcomp1}, \cref{tab:trndsann}).  Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis and are important independent of overall trends.  Nutrient concentrations were highest at the peripheral stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento, San Joaquin, Cosumnes, and Mokelumne rivers.  The highest concentrations among all nitrogen analytes, ammonium in particular, were observed at P8 as a direct consquence of \ac{wwtp} inputs upstream.  Elevated ammonium concentrations were also observed at C3 as a measure of upstream contributions from the Sacramento Regional County Sanitation District.  By contrast, nitrite/nitrate concentrations were highest at C10 as a measure of contributions from the San Joaquin River to the south that drains a predominantly agricultural watershed.  Although the Sacramento River drains a much larger area, the dominant ammonium signal as compared to nitrite/nitrate at C3 underscores the importance of \ac{wwtp} control for water quality issues downstream.

Differing magnitudes of nitrogen analytes between stations as a function of source type can have an effect on the relationship between flow and nutrients.  Both \citet{Hirsch10} and \citet{Beck15b} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.  In particular, a chemodynamic response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources \citep{Thompson11,Wan17}.  Increased flow may induce a change in nutrient concentrations, such that reduction may occur with flushing or an increase may occur through mobilization.  By contrast, nutrient loads are relatively chemostatic or invariant with changes in flow if point-sources are the dominant contributor.  These relationships are modelled particularly well with \ac{wrtds}, which can provide a means of hypothesizing unknown sources or verifying trends in response to management actions.  As noted above, C10 at the inflow of the San Joaquin River is dominated by nitrite/nitrate consistent with diffuse, agricultural inputs from the watershed.  A logical expectation is that trends from observed data may vary considerably from trends with modelled results that are flow-normalized.  Accordingly, trend analysis of nitrate/nitrate by year and season showed that percent changes at C10 were typically underestimated with the observed data during the recent period from 1996-2013 (\cref{tab:trndsann,tab:trndsaft}).  This is consistant with an expected effect of flow on raw time series, particularly for chemodynamic behavior at locations that drain highly developed watersheds \citep{Wan17}.

% obs v flow norm trends
Our results demonstrated the potential to interpret different trends if flow effects are not considered, where the difference could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions. An alternative evaluation of flow in the Delta demonstrated that flow contributions from different sources vary considerably over time at each station \citep{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.

>>>>

\section*{Abstract}
\noindent Quantitative descriptions of water quality, physical, and biological characteristics in estuaries is critical for developing an ecological understanding of drivers of change.  Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \acl{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.  Trend analysis with flow-normalized results demonstrated the potential to interpret different changes relative to observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed. Modelled trends demonstrated that nutrient concentrations were on average higher in the last twenty years relative to the earlier periods of observation, although a consistent decline began in the mid-1990s and early 2000s. We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \acl{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \acl{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and influenced by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

Understanding drivers of water quality change in estuaries depends on accurate descriptions of source inputs. The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows in the upper \ac{sfe} that receives and processes nutrient inputs from the highly agricultural watershed of the Central Valley \citep{Jassby00,Jassby02,Jassby08}.  Although water quality conditions in \ac{sfe} symptomatic of eutrophication have historically been infrequent, recent responses to stressors suggests that ecosystem condition may be changing from past norms. Increases in phytoplankton biomass, reductions in dissolved oxygen, and increasing abundance of species associated with harmful algal blooms (e.g., \textit{Pseudo-nitzschia}, \textit{Alexandrium}, \textit{Dinophysis spp.}) have been a recent concern for the management of this prominent system \citep{Lehman05,Cloern05,Cloern07,Shellenbarger08}.  As a result, \ac{chla} thresholds to assess and manage levels of concern for phytoplankton biomass in the lower estuary have been proposed \citep{Sutula17}. Although these changes are linked to drivers at different spatial and temporal scales, describing inputs from the Delta is critical to understand downstream effects.  
Rates of primary production in coastal habitats are often defined by nutrient concentrations, although a simple relationship between enrichment and water quality changes can be difficult to determine \citep{Cloern01}. Nutrient concentrations are generally non-limiting for phytoplankton growth in the upper \ac{sfe}, whereas light availability is the primary limiting factor preventing accumulation of phytoplankton biomass \citep{Cole84,Alpine88}. Grazing pressure from pelagic fishes and benthic invertebrates can also reduce phytoplankton during periods of growth \citep{Nichols85,Jassby08,Kimmerer14}.  Moreover, changes in flow management practices compounded with climate variation have altered flushing rates and turbidity as key factors that moderate phytoplankton growth in the system \citep{Alpine92,Lehman00,Wright04,Canuel09}.  \citet{Glibert14} attributed recent phytoplankton blooms in Suisun Bay to a drought, during which residence times and nitrification rates increased.  Speciation changes in the dominant forms of nitrogen are considered key factors that contribute to phytoplankton blooms, particularly seasonal reductions in ammonium that allow uptake of nitrate that stimulates growth \citep{Dortch90,Dugdale07}. Although phytoplankton concentrations have been relatively consistent in recent years in Suisan Bay, biomass in the upper Delta has been increasing \citep{Jassby08}.  Much of these trends were explained by invasion of benthic grazers in polyhaline areas and changes in the mean flow conditions observed in the Delta.  Descriptions of nitrogen trends over several decades could be used to understand these recent changes, particularly in the context of primary production and physical drivers of change.

Long-term monitoring data are powerful sources of information that can facilitate descriptions of water quality change in coastal regions. A comprehensive water quality monitoring program has been in place in the Delta for several decades \citep[\cref{fig:delt_map},][]{IEP13}.  Although these data have been used extensively \citep[e.g.,][]{Lehman92,Jassby08,Glibert10}, water quality trends covering the full spatial and temporal coverage of the monitoring dataset have not been systematically evaluated. Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.  An evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that the majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported to the estuary \citep{Novick15}. Seasonal and annual changes in the delivery of water inflows and water exports directly from the system can also obscure trends \citep{Jassby00,Jassby08}.  A complete assessment of the Delta dataset that considers these variable effects is necessary to characterize the different trends in nitrogen analytes.

@@

The goal of this study was to provide a comprehensive description of nutrient trends in the Delta over the last forty years.  This information can inform the  understanding of ecosystem response dynamics and potential causes of water quality change. The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
\item provide detailed descriptions of two case studies in the context of conceptual relationships modeled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production.  Our general hypothesis was that the results were expected to support previous descriptions of trends in this well-studied system, but that new insight into spatial and temporal variation in response endpoints was expected, particuarly in flow-normalized model predictions.

@@

Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.  Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.  Stations were grouped by location in the study area for comparison: \textit{peripheral} Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; \textit{interior} Delta stations D19, D26, D28; and \textit{Suisun} stations D4, D6, and D7.  These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}. Although many other stations are available for the region, the stations were chosen because they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}. Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual ($\beta_1$) or seasonal ($\beta_3$, $\beta_4$) trend, and $Q$ is the flow variable (either flow or salinity depending on station).

The \ac{wrtds} model is a moving window regression that fits unique parameters (i.e., $\beta_0$, \ldots, $\beta_4$) at each observation point in the time series ($n$ ranging from \Sexpr{minobs} at \Sexpr{minsta} to \Sexpr{maxobs} at \Sexpr{maxsta}).  Rather than fitting a global model to the entire time series, one regression is fit to every observation.  Observations within a window for each regression are weighted relative to annual, seasonal, and flow distances from the observation at the center of the window.  Observations with distances farther from the center (i.e., greater time and different flow values from the center) have less weight during parameter estimation for each regression.  This approach allows for a type of smoothing where the observed fit is specific to the data characteristics within windows. Models applied herein were based on a tidal adaptation of the original method that can use either flow or salinity estimates as nutrient predictors \citep{Beck15}.  All models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data below the detection limit \citep{Tobin58}.  Model predictions were evaluated as monthly values or as annual values that averaged monthly results within each water year (October to September). All analyses used the WRTDStidal package for the R statistical programming language \citep{Beck17b,RDCT17}.  The default model fitting procedures were used that set half-window widths as six months for seasonal weights, ten years for annual weights, and half the range of salinity or flow in the input data for $Q$ weights.

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows \citep{Hirsch10} or tidal variation \citep{Beck15}. Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods. Flow-normalization predictions for each month of each year were based on the average of predictions for flow values that occur in the same month across all years, weighted within each specific month and year for every observation. Flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping. Annual groupings were chosen as approximate twenty year midpoints in the time series and seasonal groupings were chosen to evaluate inter-annual changes while keeping season constant.

@@

\subsection{Observed data and modelled trends}

@@

Relative to the observed data, long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).  Although each station varied in the overall concentrations, patterns within the three Delta regions (peripheral, interior, and Suisun) were observed.  Concentrations for all nitrogen analytes were highest in the peripheral stations.  Ammonium concentrations at P8 and C3 were highest and showed a consistent increase over time, followed by a reduction beginning in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.  By contrast, \ac{din} and nitrite/nitrate concentrations at the peripheral stations showed increases at  P8 and C10 followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did not show any noticeable trends.  Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for D26.  Trends in \ac{din} and nitrite/nitrate for the interior stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.  These trends were similar for the Suisun stations, although the reduction in the late 1990s did not occur.  By contrast, ammonium concentrations were low in Suisun but a gradual increase over the period of record was observed.

@@

@
A comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends. For all \Sexpr{eng(nrow(zcmp))} trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period (annual or seasonal aggregations), \Sexpr{eng(sum(zcmp$nstosig))} comparisons had trends that were insignificant ($p > 0.05$) with the observed data but significant with flow-normalized results, whereas only \Sexpr{eng(sum(zcmp$sigtons))} trend changed to insignificant. This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that \Sexpr{eng(sum(chgtb[c('morepos', 'moreneg')]))} comparisons showed an increase in magnitude (more negative or more positive) and \Sexpr{eng(sum(chgtb[c('lesspos', 'lessneg')]))} had a decrease (less positive or less negative) compared to observed trends.  \Sexpr{cap(eng(chgtb['postoneg']))} comparisons showed a trend reversal from positive to negative estimated change, \Sexpr{eng(chgtb['zerotoneg'])} sites went from no change to negative estimated change, and \Sexpr{eng(chgtb['zerotopos'])} site went from no change to a positive trend for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}). The most notable change in the flow-normalized results was an overall decrease (less positive trend or greater negative trend) in concentrations for most sites in the summer and fall seasons for 1996-2013.  More statistically significant trends were also observed with the flow-normalized results.

@@

Water quality conditions in the Delta are dynamic and not easily characterized from observed time series.  Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions of the Delta that is not possible with raw observations.  A general conclusion is that nitrogen concentrations have showed a consistent decrease beginning in the mid-1990s and early 2000s, although average concentrations remain above those observed earlier in the period of record.  These results are confirmed visually from \ac{wrtds} (\cref{fig:prddat}) and through significant results from trend tests (\cref{fig:trndcomp1}, \cref{tab:trndsann}).  Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis and are important independent of overall trends.  Nutrient concentrations were highest at the peripheral stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento, San Joaquin, Cosumnes, and Mokelumne rivers.  The highest concentrations among all nitrogen analytes, ammonium in particular, were observed at P8 as a direct consquence of \ac{wwtp} inputs upstream.  Elevated ammonium concentrations were also observed at C3 as a measure of upstream contributions from the Sacramento Regional County Sanitation District.  By contrast, nitrite/nitrate concentrations were highest at C10 as a measure of contributions from the San Joaquin River to the south that drains a predominantly agricultural watershed.  Although the Sacramento River drains a much larger area, the dominant ammonium signal as compared to nitrite/nitrate at C3 underscores the importance of \ac{wwtp} control for water quality issues downstream.

Differing magnitudes of nitrogen analytes between stations as a function of source type can have an effect on the relationship between flow and nutrients.  Both \citet{Hirsch10} and \citet{Beck15} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.  In particular, a chemodynamic response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources \citep{Thompson11,Wan17}.  Increased flow may induce a change in nutrient concentrations, such that reduction may occur with flushing or an increase may occur through mobilization.  By contrast, nutrient loads are relatively chemostatic or invariant with changes in flow if point-sources are the dominant contributor.  These relationships are modelled particularly well with \ac{wrtds}, which can provide a means of hypothesizing unknown sources or verifying trends in response to management actions.  As noted above, C10 at the inflow of the San Joaquin River is dominated by nitrite/nitrate consistent with diffuse, agricultural inputs from the watershed.  A logical expectation is that trends from observed data may vary considerably from trends with modelled results that are flow-normalized.  Accordingly, trend analysis of nitrate/nitrate by year and season showed that percent changes at C10 were typically underestimated with the observed data during the recent period from 1996-2013 (\cref{tab:trndsann,tab:trndsaft}).  This is consistant with an expected effect of flow on raw time series, particularly for chemodynamic behavior at locations that drain highly developed watersheds \citep{Wan17}.

% obs v flow norm trends
Our results demonstrated the potential to interpret different trends if flow effects are not considered, where the difference could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions. An alternative evaluation of flow in the Delta demonstrated that flow contributions from different sources vary considerably over time at each station \citep{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.  Our analysis is the first attempt to model nutrient dynamics related to flow in the entire Delta, such that additional work should focus on improving the characterization of flow signals at each station.
